Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-10-27 src/tex/ms.tex
[showyourwork] overleaf sync

    \begin{figure*}
        \centering\includegraphics{figures/comparison-scenarios.pdf}
        \caption{Illustration of the expected counts for the different source and instrument scenarios we used to make test datasets. The source patterns shown are for the {\sl (A)}  {\tt points}, {\sl (B) {\tt asterism}}, {\sl (C)} {\tt shield}, and {\sl (D)} {\tt spiral} cases (see Section~\ref{sec:testmodels}). For the instrument scenario \chandra we convolved the with a narrow Gaussian shaped PSF of width $\sigma=2$~pix. For the scenario \xmm we used a Gaussian shaped PSF with a width of $\sigma=2$~pix and a five times increased effective are. All images are of size $128{\times}128$ pixels. \vlk{Please avoid using "A.U." for anything but Astronomical Units!}
        }
        \label{fig:scenarios}
        \script{comparison-scenarios.py}
    \end{figure*}

@@

    For scenario B3 all methods recover the point sources well. However, a large difference can be observed in the reconstruction of the Gaussian shaped extended sources. As expected, the uniform prior assumption lets the emission decompose into many point sources, neglecting the spatial correlation between them. Pylira shows slightly improved results with the gaps between the bright emission peeks being more filled, however, it still lacks a convincing result of reconstructing the source as a single object. The \jolideco method with the patch priors shows the best results both visually as well as measured by the SSI. The GLEAM v0.1 patch prior reconstructs the sources almost perfectly, visually almost indistinguishable from the ground truth. The Zoran \& Weiss prior also shows good reconstruction of the extended sources, but less smooth and leaving a more visible extension in the surrounding point sources.

@@

    \subsubsection{Flux Profile}
    To get a more qualitative measurement of the improvement we compute a flux profile on the image of E0102 reconstructed by \jolideco. For this we chose a region perpendicular to the filament shown in Zoom A of Figure~\ref{fig:example-chandra}. To estimate the errors we use the bootstrapping method. For each individual observation we sample from the event list with replacement. For each realization of the set of observation we run \jolideco and compute a flux profile on the reconstructed image. This way we obtain a distribution of profiles, which we can use to compute an error band. For comparison we also compute the profiles and associated errors directly on the stacked bootstrapped images of counts. The result is shown in Figure~\ref{fig:chandra-e0102-zoom-a}.

    \vlk{We compute the full-width at half-max (FWHM) of each of the profiles, for both the original counts and the deconvolved intensities, and find that the deconvolved profile is narrower.  For the counts profile, FWHM$_C=0.63 \pm 0.09$~arcsec, where the error bar is obtained by computing the standard deviations over the bootstrap sample; and the deconvolved intensity profile has FWHM$_J=0.56 \pm 0.03$~arcsec.  Note that the PSF itself has a FWHM$_{{\rm PSF}\approx{0.6}}$~arcsec\footnote{See projected radial profiles of the on-axis PSF at \url{https://space.mit.edu/cxc/marx/tests/PSF.html}.  The 1$\sigma$ radius is $\approx$0.26~arcsec (corresponding to 39\% enclosed counts in 2-D), suggesting that FWHM$_{{\rm PSF}}\approx{2.355{\cdot}0.26=0.62}$~arcsec.}.  }
    The profile computed from the images reconstructed with \jolideco has a much smaller noise level and the peak is much more pronounced. By fitting the profile with a univariate spline and finding its roots we measure the full width half maximum of the curves. We find $w_J=0.56\pm0.03$ for the \jolideco profile and $w_C=0.63\pm0.09$ for the profile computed on the counts image. Assuming that the PSF and the shape of peak of the profile are approximately Gaussian a we expect the following relation between the width of the counts profile, the PSF and the flux profile:

    \begin{equation}
        w_{J} = \sqrt{w_C^2 - \w_{\mathrm{PSF}}^2}
    \end{equation}

    With $w_{\mathrm{PSF}} \approx 0.25 \mathrm{arcsec}$ \footnote{The 38\% containment radius of the Chandra PSF is $\sigma_{39}=2$ native pixels. This corresponds to  ee e.g. \url{https://space.mit.edu/cxc/marx/tests/PSF.html}}
    we obtain ..., which is consistent with the values we measured for the profile. The profile computed from the images reconstructed with \jolideco show both a smaller width of the profile as well as a much smaller error.

>>>>

    \begin{figure*}
        \centering\includegraphics{figures/comparison-scenarios.pdf}
        \caption{Illustration of the expected counts for the different source scenarios we used to make test datasets. The source patterns shown are for the {\sl (A)}  {\tt points}, {\sl (B) {\tt asterism}}, {\sl (C)} {\tt shield}, and {\sl (D)} {\tt spiral} cases (see Section~\ref{sec:testmodels}). For the instrument scenario \chandra we convolved the with a narrow Gaussian shaped PSF of width $\sigma=2$~pix. For the scenario \xmm we used a Gaussian shaped PSF with a width of $\sigma=2$~pix and a five times increased effective are. All images are of size $128{\times}128$ pixels. \vlk{Please avoid using "A.U." for anything but Astronomical Units!}
        }
        \label{fig:scenarios}
        \script{comparison-scenarios.py}
    \end{figure*}

@@

    For scenario B3 all methods recover the point sources well. However, a large difference can be observed in the reconstruction of the Gaussian shaped extended sources. As expected, the uniform prior assumption lets the emission decompose into many point sources, neglecting the spatial correlation between them. Pylira shows slightly improved results with the gaps between the bright emission peeks being more filled, however, it still lacks a convincing result of reconstructing the source as a single object. The \jolideco method with the patch priors shows the best results both visually as well as measured by the SSI. With the GLEAM v0.1 patch prior the reconstruction is almost perfectly, visually almost indistinguishable from the ground truth. The Zoran \& Weiss prior also shows good reconstruction of the extended sources, but less smooth and leaving a more visible extension in the surrounding point sources.

@@

    \subsubsection{Flux Profile}
    To get a more quantitative measurement of the improvement in angular resolution we compute a flux profile on the image of E0102 reconstructed by \jolideco. For this we chose a region perpendicular to the filament shown in Zoom A of Figure~\ref{fig:example-chandra}. To estimate the errors we use the bootstrapping method. For each individual observation we sample from the event list with replacement. For each realization of the set of observation we run \jolideco and compute a flux profile on the reconstructed image. This way we obtain a distribution of profiles, which we can use to compute an error band. For comparison we also compute the profiles and associated errors directly on the stacked bootstrapped images of counts. The result is shown in Figure~\ref{fig:chandra-e0102-zoom-a}.

    We compute the full-width at half-max (FWHM) of each of the profiles, by fitting a Gaussian shape with a linear gradient as baseline 
    for both the original counts and the deconvolved intensities. We find that the deconvolved profile is significantly narrower. For the counts profile, $\omega_C=0.77 \pm 0.22$~arcsec, where the error bar is obtained by computing the standard deviations over the bootstrap sample; and the deconvolved intensity profile has $\omega_J=0.52 \pm 0.03$~arcsec.

    Assuming that the PSF and the shape of peak of the profile are approximately Gaussian we expect the following relation between the width of the counts profile, the PSF and the flux profile:
    %
    \begin{equation}
        \omega_{J} = \sqrt{\omega_C^2 - \omega_{\mathrm{PSF}}^2}
    \end{equation}
    %
    With $w_{\mathrm{PSF}} \approx 0.6 \mathrm{arcsec}$ \footnote{See projected radial profiles of the on-axis PSF at \url{https://space.mit.edu/cxc/marx/tests/PSF.html}.  The 1$\sigma$ radius is $\approx$0.26~arcsec (corresponding to 39\% enclosed counts in 2-D), suggesting that FWHM$_{{\rm PSF}}\approx{2.355{\cdot}0.26=0.62}$~arcsec.}, we expect a FWHM after de-convolution of $\approx\omega_J=0.48$~arcsec. This is roughly consistent with the values we measured for the profile.
